### 2016-05-03 paper/icfp16-submission/rebuttal.txt

We thank the reviewers for their thoughtful remarks, and will respond
briefly to a few categories of concerns.

1. Related Work

We thank reviewer C for the pointers to extra related work. We will
greatly expand the section in the next revision of the paper.

TODO: list a few additional citations we would talk about

2. Bias in Type Variable Instantiation

In the `fac` example, asserting that `n` must be an int without looking
at the rest of the code is indeed biased, but any inference scheme will
be biased in one direction or another. Our bias is towards the
evaluation semantics of the language, other techniques are biased by
their traversal strategy (eg top-down vs bottom-up) or by various
heuristics (whether provided by the compiler or through generic
reasoning).

We would argue, however, (as in Section 5.4) that our bias is more
natural as it is based on a concrete instance of the program crashing.

3. Overloading / Type Classes

Overloaded operators can be problematic for our approach. In reviewer
C's variant of `fac`, our approach would get stuck at the `if n <= m`
test, because we cannot know how to instantiate `n` or `m` at that point
in the program. This does not signify a type error of course, but our
approach would fail to discover a witness. We suspect this is relatively
uncommon in novice OCaml programs (TODO: there were no instances of
this scenario in either of our datasets), but the issue would surely be
exacerbated if we ported the approach to Haskell due to the prevalence
of type classes.

We will add a paragraph describing this issue to the Discussion section,
and defer a treatment of overloading to future work.

4. Scaling to Real-World Code

Our benchmarks are small programs as they were drawn from homework
submissions. The UW dataset had programs generally in the 100-200 LoC
range, while the programs from our dataset were generally <50 LoC (we
had already sliced out a minimal program for each error in our dataset).
It is quite possible that our heuristic for compressing the trace based
on function calls would not suffice for larger programs, though there is
a broad literature on program slicing to draw from there.

5. Language Features

By "pure subset" of OCaml we mean polymorphic datatypes and records, and
we will clarify this in the text. We do not support references, modules,
or the object system.

It is true that we implicitly assume all types are inhabited in the
formalism, and that the generality theorem is weaker in the presence of
uninhabited types. We will clarify this in the text.

### 2016-05-03 paper/icfp16-submission/rebuttal.txt

In the `fac` example, asserting that `n` must be an int without looking
at the rest of the code is indeed biased, but any inference scheme will
be biased in one direction or another. Our bias is towards the
evaluation semantics of the language, other techniques are biased by
their traversal strategy (eg top-down vs bottom-up) or by various
heuristics (whether provided by the compiler or through generic
reasoning).

We would argue, however, (as in Section 5.4) that our bias is more
natural as it is based on a concrete instance of the program crashing.

3. Overloading / Type Classes

Overloaded operators can be problematic for our approach. In reviewer
C's variant of `fac`, our approach would get stuck at the `if n <= m`
test, because `n` and `m` are both unconstrained at this point (apart
from having the same type), so we do not know how to instantiate
them. This does not signify a type error of course, but our approach
would fail to discover a witness. We suspect this is relatively uncommon
in novice OCaml programs (there were no instances of this scenario in
either of our datasets), but the issue would surely be exacerbated if we
ported the approach to Haskell due to the prevalence of type classes.

### 2016-05-04 paper/icfp16-submission/rebuttal.txt

We thank the reviewers for their thoughtful remarks, and will respond
to a few categories of concerns.

@@

1. Related Work

We thank reviewers B and C for the pointers to extra related work. We
will greatly expand the section in the next revision of the paper to
include many of the following references.

- Haack and Wells, ESOP 2003
- Rahli et al., Tech Report 2010
- Schilling, TFP 2011
- Gast, IFL 2004
- Heeren et al., ICFP 2003
- Hage and Heeren, IFL 2006
- Christiansen, TFP 2014
- Serrano and Hage, ESOP 2016
- Chitil, ICFP 2001
- Stuckey et al., Haskell 2003
- Bayne et al., ICSE 2011
- Zhang et al., PLDI 2015

### 2016-05-04 paper/icfp16-submission/rebuttal.txt

We thank the reviewers for their thoughtful remarks, and will respond
to a few categories of concerns.

1. Related Work

We thank reviewers B and C for the pointers to extra related work.

Type error slicing shares a common goal with us of providing
more diagnostic information to the programmer, though the techniques
differ significantly. Error slicing is based on a static analysis of
the typing data, usually as a constraint system, and presents
multiple candidate error locations, possibly ranked according to
some heuristic. In contrast, our approach executes the program and
presents a concrete execution trace demonstrating the error. Gast's
IFL 2004 paper is perhaps the most similar to ours as he produces
a data flow that would violate the typing rules.

Customizable type error messages along the lines of Heeren 2003,
Christiansen 2015, and Serrano Mena 2016 also addresses the issue
of poor type errors. The difference with our work is more pronounced,
as these techniques often rely on a library author (or instructor)
with domain-specific knowledge to craft high-quality error messages,
whereas our approach is fully automatic but ignorant of the application
domain.

We might also add the algorithmic type debugging work of Chitil 2001
and Stuckey 2003, which presents the user with a series of queries
about the types of intermediate expressions to help isolate and
explain the error.

We will extend the Related Work section to reference all of these
techniques and papers.

### 2016-05-04 paper/icfp16-submission/rebuttal.txt

We thank the reviewers for their thoughtful remarks, and will respond
to a few categories of concerns.

1. Related Work

We thank reviewers B and C for the pointers to extra related work.

We share a common goal with type error slicing, though the techniques
differ significantly. Error slicing involves a static analysis of the
typing data and presents multiple candidate error locations, possibly
ranked according to some heuristic. In contrast, we execute the program
and present an execution trace demonstrating the error. Gast's IFL 2004
paper is perhaps the most similar to ours as he produces a data flow
that would violate the typing rules.

Customizable type error messages along the lines of Heeren 2003,
Christiansen 2015, and Serrano Mena 2016 also addresses the issue of
poor type errors. The difference with our work is more pronounced, as
these techniques often rely on domain-specific knowledge to craft
high-quality error messages, whereas our approach is fully automatic but
ignorant of the application domain.

We will extend the Related Work section to reference all of these
techniques and papers.

2. Bias in Type Instantiation

In the `fac` example, asserting that `n` must be an int at the branch is
indeed biased. Our bias is towards the evaluation semantics of the
language, other techniques are biased by their traversal strategy or by
various heuristics.

We would argue, however, (as in Section 5.4) that our bias is more
natural as it is based on a concrete instance of the program crashing.
We will expand the relevant text to clarify this issue.

@@

3. Overloading / Type Classes

Overloaded operators are problematic for our approach. In reviewer C's
variant of `fac`, we would get stuck at the `if n <= m` test because `n`
and `m` are constrained to the same unknown type. Thus we do not know
how to instantiate them, and would fail to discover a witness. Our
benchmarks suggest this is rare in novice programs, but the issue would
be exacerbated if we ported the approach to Haskell, due to the
prevalence of type classes.

We will describe this issue in the Discussion section, and defer a
treatment of overloading to future work.

@@

4. Scaling to Real-World Code

Our benchmarks are small programs as they were drawn from homework
submissions, the largest being around 200 LoC. It is quite possible
that our heuristic for compressing the trace based on function calls
would not suffice for larger programs, though there is a broad
literature on program slicing to draw from there. We will augment our
discussion of this threat to generality in the text.

### 2016-05-04 paper/icfp16-submission/rebuttal.txt

Thank you for your thoughtful feedback!

1. Related Work

We thank reviewers B and C for the pointers to extra related 
work -- we will extend the Related Work section to reference 
all of these techniques and papers.

@@

In the `fac` example, asserting that `n` must be an int at the branch is
indeed biased. Our bias is towards the evaluation semantics of the
language, other techniques are biased by their traversal strategy or by
various heuristics.

We would argue, however, (as in Section 5.4) that our bias is more
natural as it is based on a concrete instance of the program crashing.
We will expand the relevant text to clarify this issue.

3. Overloading / Type Classes

Currently our approach does not support overloading.

With an overloaded language, reviewer C's variant of `fac`, 
we would get stuck at the `if n <= m` test because `n` and `m` 
are constrained to the same unknown type, but as we do not know 
how to instantiate such types, we  would fail to discover a witness.

Thus, our formalization (theory) assumes there are no overloaded
(or polymorphic) operators. Furthermore, in our implementation, 
i.e. over *all* our benchmarks,  all the operands are concretized 
*before* they are applied the any (overloaded) operator.

@@

We will describe this issue in the Discussion section, and defer a
treatment of overloading to future work.

4. Scaling to Real-World Code

Our benchmarks are small programs as they were drawn from homework
submissions, the largest being around 200 LoC. While our jump compression 
heuristic may not suffice for larger programs, we could extend it with 
ideas from the slicing literature.

@@

5. Language Features

By "pure subset" of OCaml we mean polymorphic datatypes and records, 
and we will clarify this in the text. Currently, we do not support 
references, modules, or the object system.

Indeed, we implicitly assume all types are inhabited in the formalism, 
and that the generality theorem is weaker in the presence of uninhabited 
types. We will clarify this in the text.

### 2016-05-04 paper/icfp16-submission/rebuttal.txt

We thank reviewers B and C for the pointers to extra related 
work -- we will extend the Related Work section to reference 
all of these techniques and papers.

We share a common goal with type error slicing, though the techniques
differ significantly. Error slicing involves a static analysis of the
typing data and presents multiple candidate error locations, possibly
ranked according to some heuristic. In contrast, we execute the program
and present an execution trace demonstrating the error. Gast's IFL 2004
paper is perhaps the most similar to ours as he produces a data flow
that would violate the typing rules.

Customizable type error messages along the lines of Heeren 2003,
Christiansen 2015, and Serrano Mena 2016 also addresses the issue of
poor type errors. The difference with our work is more pronounced, as
these techniques often rely on domain-specific knowledge to craft
high-quality error messages, whereas our approach is fully automatic but
ignorant of the application domain.

2. Bias in Type Instantiation [C]

Our goal is to be as lenient as possible in type instantation. In the
`fac` example, we assert that `n` must be an int at the branch condition
because that is the only way to make progress. This is indeed biased, as
the `n <= 0` expression may be the source of the error, but it is
necessary to continue evaluation.

We will expand the relevant text to clarify this issue.

3. Overloading / Type Classes [A,B,C]

### 2016-05-24 paper/icfp16-cameraready/evaluation.tex
misc

\paragraph{Benchmarks}
We answer these questions on two sets of ill-typed programs, \ie
programs that were rejected by the \ocaml\ compiler because of a
type error.
%
The first dataset comes from the Spring 2014 undergraduate Programming
Languages course at our institution.
%
We recorded each interaction with the \ocaml\ top-level system over the
course of the first three assignments (IRB  
\# hidden for blind review), 
% \#140608),
from which we extracted \ucsdsize\ distinct, ill-typed \ocaml programs.
%
The second dataset --- widely used in the literature --- comes from a
similar course at the University of Washington~\cite{lerner_seminal:_2006},
from which we extracted 284 ill-typed programs.
%
Both datasets include programs that demonstrate a variety of functional
programming idioms, including (tail) recursive functions, higher-order
functions, polymorphic data types, and expression evaluators.

@@

\begin{figure*}[t]
\centering
\begin{minipage}{\linewidth}
\centering
\includegraphics[width=0.49\linewidth]{coverage.png}
% \end{minipage}
% \begin{minipage}{\linewidth}
\includegraphics[width=0.49\linewidth]{distrib.png}
\end{minipage}
% \vspace{-8ex}
\caption{Results of our coverage testing and the distribution of test
  outcomes. Our random search successfully finds witnesses for 83\% of
  the programs in under one second, improving slightly to 87\% in under
  10 seconds. In both datasets we detect actual type errors about 82\%
  of the time, unbound variables or constructors 3-4\% of the time, and
  diverging loops 2-3\% of the time. For the remaining 11-12\% of the
  programs we are unable to provide any useful feedback.  }
\label{fig:results-witness}
\end{figure*}

\paragraph{Results}
\label{sec:results-witness}
The results of our experiments are summarized in
Figure~\ref{fig:results-witness}.
%
In both datasets our tool was able to find a witness for 83\% of the
programs in under one second, \ie fast enough to be integrated as a
compile-time check. If we extend our tolerance to a 10 second timeout,
we hit a maximum of 87\% coverage.
%
Interestingly, while the vast majority of witnesses corresponded to a
type-error, as expected, 3-4\% triggered an unbound variable error (even
though \ocaml reported a type error) and 2-3\% triggered an infinite
recursion error.
%
For the remaining 12\% of programs we were unable to provide any useful
feedback as they either completed 1,000 tests successfully, or timed out
after 10 minutes.
%
% XX programs were deemed safe and XX timed out even at 3,000 steps, \ie
% we could not provide any useful feedback for XX\% of the total programs.
%
While a more advanced search procedure, \eg dynamic-symbolic execution,
could likely trigger more of the type errors, our experiments show that
type errors are coarse enough (or that novice programs are \emph{simple}
enough) that these techniques are not necessary.

@@

\paragraph{Trace Explosion}
Though the average complexity of our generated traces is low in terms of
jumps, there are some extreme outliers. We cannot reasonably expect a
novice user to explore a trace containing 50+ terms and draw a
conclusion about which pieces contributed to the bug in their
program. Enhancing our visualization to slice out program paths relevant
to specific values~\cite{perera_functional_2012},
would likely help alleviate this issue, allowing users to
highlight a confusing value and ask: ``Where did this come from?''

>>>>

\paragraph{Benchmarks}
We answer these questions on two sets of ill-typed programs, \ie
programs that were rejected by the \ocaml\ compiler because of a
type error.
%
The first dataset comes from the Spring 2014 undergraduate Programming
Languages course at our institution.
%
We recorded each interaction with the \ocaml\ top-level system over the
course of the first three assignments (IRB  
\# hidden for blind review), 
% \#140608),
from which we extracted \ucsdsize\ distinct, ill-typed \ocaml programs.
%
The second dataset --- widely used in the literature --- comes from a
graduate-level course at the University of Washington~\cite{Lerner2006-pj},
from which we extracted 284 ill-typed programs.
%
Both datasets include programs that demonstrate a variety of functional
programming idioms, including (tail) recursive functions, higher-order
functions, polymorphic data types, and expression evaluators.

@@

\begin{figure*}[t]
\centering
\begin{minipage}{\linewidth}
\centering
\includegraphics[width=0.49\linewidth]{coverage.png}
% \end{minipage}
% \begin{minipage}{\linewidth}
\includegraphics[width=0.49\linewidth]{distrib.png}
\end{minipage}
% \vspace{-8ex}
\caption{Results of our coverage testing and the distribution of test
  outcomes. Our random search successfully finds witnesses for 83\% of
  the programs in under one second, improving slightly to 87\% in under
  10 seconds. In both datasets we detect actual type errors about 82\%
  of the time, unbound variables or constructors 3--4\% of the time, and
  diverging loops 2--3\% of the time. For the remaining 11--12\% of the
  programs we are unable to provide any useful feedback.  }
\label{fig:results-witness}
\end{figure*}

\paragraph{Results}
\label{sec:results-witness}
The results of our experiments are summarized in
Figure~\ref{fig:results-witness}.
%
In both datasets our tool was able to find a witness for 83\% of the
programs in under one second, \ie\ fast enough to be integrated as a
compile-time check. If we extend our tolerance to a 10 second timeout,
we hit a maximum of 87\% coverage.
%
Interestingly, while the vast majority of witnesses corresponded to a
type-error, as expected, 3--4\% triggered an unbound variable error (even
though \ocaml\ reported a type error) and 2--3\% triggered an infinite
recursion error.
%
For the remaining 12\% of programs we were unable to provide any useful
feedback as they either completed 1,000 tests successfully, or timed out
after 10 minutes.
%
% XX programs were deemed safe and XX timed out even at 3,000 steps, \ie
% we could not provide any useful feedback for XX\% of the total programs.
%
While a more advanced search procedure, \eg\ dynamic-symbolic execution,
could likely trigger more of the type errors, our experiments show that
type errors are coarse enough (or that novice programs are \emph{simple}
enough) that these techniques are not necessary.

@@

\paragraph{Trace Explosion}
Though the average complexity of our generated traces is low in terms of
jumps, there are some extreme outliers. We cannot reasonably expect a
novice user to explore a trace containing 50+ terms and draw a
conclusion about which pieces contributed to the bug in their
program. Enhancing our visualization to slice out program paths relevant
to specific values~\cite{Perera2012-dy},
would likely help alleviate this issue, allowing users to
highlight a confusing value and ask: ``Where did this come from?''

### 2016-05-24 paper/icfp16-cameraready/gallery2.tex
misc

\subsection{The Advantage Of Traces}\label{sec:advantage-traces}

Next, we present a \emph{qualitative} evaluation that compares
the explanations provided by \toolname's dynamic witnesses with
the static reports produced by the \ocaml\ compiler and \sherrloc,
a state-of-the-art fault localization approach~\cite{zhang_toward_2014}.
%
In particular, we illustrate, using a series of examples drawn
from student programs in the \ucsdbench\ dataset, how \toolname's
jump-compressed traces can get to the heart of the error. Our approach
%
highlights the conflicting values that cause the program to get
stuck, rather that blaming a single one,
%
shows the steps necessary to reach the stuck state, and
%
does not assume that a function is correct just because it type-checks.
%
For each example we will present
(1)~the code,
(2)~the error message returned \ocaml,
(3)~the error locations returned by \ocaml\ (underlined) and \sherrloc\ (in bold),\footnote{When the locations from \ocaml\ and \sherrloc\ overlap,
we just underline the relevant code.}
and (4)~the jump-compressed trace produced by \toolname.

@@

  let rec digitsOfInt n =
    if n <= 0 then
      []
    else
      append (==digitsOfInt (n / 10)==)
             [__n mod 10__]
\end{ecode}
%
Unfortunately, the student's @append@ function \emph{conses} an element
onto a list instead of appending two lists.
%
Though incorrect, @append@ still type-checks and thus \ocaml and
\sherrloc blame the \emph{use-site} on line 10.
%
\begin{verbatim}
  This expression has type
    int
  but an expression was expected of type
    'a list
\end{verbatim}
%
\toolname, in contrast, makes no asummptions about @append@ and produces
a trace that illustrates the true error on line 4, by
highlighting the conflict in consing a list onto a list of integers.
%
\begin{center}
  \includegraphics[height=175px]{digitsOfInt.png}
\end{center}
%

>>>>

\subsection{The Advantage of Traces}\label{sec:advantage-traces}

Next, we present a \emph{qualitative} evaluation that compares
the explanations provided by \toolname's dynamic witnesses with
the static reports produced by the \ocaml\ compiler and \sherrloc,
a state-of-the-art fault localization approach~\cite{Zhang2014-lv}.
%
In particular, we illustrate, using a series of examples drawn
from student programs in the \ucsdbench\ dataset, how \toolname's
jump-compressed traces can get to the heart of the error. Our approach
%
highlights the conflicting values that cause the program to get
stuck, rather that blaming a single one,
%
shows the steps necessary to reach the stuck state, and
%
does not assume that a function is correct just because it type-checks.
%
For each example we will present
(1)~the code,
(2)~the error message returned \ocaml,
(3)~the error locations returned by \ocaml\ (underlined) and \sherrloc\ (in bold),\footnote{When the locations from \ocaml\ and \sherrloc\ overlap,
we just underline the relevant code.}
and (4)~the jump-compressed trace produced by \toolname.

@@

  let rec digitsOfInt n =
    if n <= 0 then
      []
    else
      append (==digitsOfInt (n / 10)==)
             [__n mod 10__]
\end{ecode}
%
Unfortunately, the student's @append@ function \emph{conses} an element
onto a list instead of appending two lists.
%
Though incorrect, @append@ still type-checks and thus \ocaml and
\sherrloc blame the \emph{use-site} on line 10.
%
\begin{verbatim}
  This expression has type
    int
  but an expression was expected of type
    'a list
\end{verbatim}
%
\toolname, in contrast, makes no assumptions about @append@ and produces
a trace that illustrates the true error on line 4, by
highlighting the conflict in consing a list onto a list of integers.
%
\begin{center}
  \includegraphics[height=175px]{digitsOfInt.png}
\end{center}
%

### 2016-05-24 paper/icfp16-cameraready/haskellListings.tex
init paper/icfp16-cameraready
misc

\lstnewenvironment{ecode}
{\lstset{language=HaskellUlisses, numbers=left, frame=leftline, moredelim=[is][\underbar]{__}{__}}}
{}

>>>>

\lstnewenvironment{ecode}
{\lstset{language=HaskellUlisses
				,numbers=left
                                ,frame=leftline
                                ,xleftmargin=7.5mm
				,moredelim=[is][\textbf]{==}{==}
				,moredelim=[is][\underbar]{__}{__}
				}}
{}

### 2016-05-24 paper/icfp16-cameraready/introduction.tex
misc

Type errors are a common stumbling block for students
trying to learn typed functional languages like \ocaml\
and \haskell.
%
Consider the ill-typed @fac@ function on the left in
Figure~\ref{fig:factorial}.
%
The function returns @true@ in the base case (instead of @1@),
and so \ocaml responds with the error message:
%
\begin{verbatim}
 File "fac.ml", line 5, characters 8-17:
 Error: This expression has type bool
        but an expression was expected
        of type int
\end{verbatim}
%
This message makes perfect sense to an expert who is familiar
with the language and has a good mental model of how the type
system works.
%
However, it may perplex a novice who has yet to develop such a
mental model.
%
To make matters worse, unification-based type inference algorithms
often report errors far removed from their source.
%
This further increases the novice's confusion and can actively mislead
them to focus their investigation on an irrelevant piece of code.
%
Much recent work has focused on analyzing unification constraints
to properly \emph{localize} a type error~\cite{lerner_searching_2007,chen_counter-factual_2014,zhang_toward_2014,pavlinovic_finding_2014},
but an accurate source location still does not explain \emph{why} the
program is wrong.

\begin{figure}[t]
\centering
\begin{minipage}{.49\linewidth}
\centering
\begin{code}
  let rec fac n =
    if n <= 0 then
      true
    else
      n * fac (n-1)
\end{code}
\vspace{3em}
% \end{minipage}
% % \hspace{0.3in}
% \begin{minipage}{.39\linewidth}
\centering
  \includegraphics[height=1.5in]{fac-overview.png}
\end{minipage}
\begin{minipage}{.49\linewidth}
\centering
\includegraphics[height=3in]{fac-long.png}
\end{minipage}
\caption{(TL) An ill-typed \texttt{fac} function; (BL) Dynamically witnessing the type error in \texttt{fac}, showing only function calls; (R) The same trace, fully expanded.}
\label{fig:factorial}
\end{figure}

@@

\paragraph{3. Evaluating Witnesses}
%
Of course, the problem of finding witnesses is
undecidable. In fact, due to the necessarily
conservative nature of static typing, there
may not even exist any witnesses for a given
ill-typed program.
%
Thus, our approach is a heuristic that is only useful
if it can find \emph{compact} witnesses for
\emph{real-world} programs.
%
Our third contribution is an extensive evaluation of our approach
on two different sets of ill-typed programs obtained by instrumenting
compilers used in beginner's classes~(\S~\ref{sec:evaluation}).
%
The first is the \uwbench\ data set~\cite{lerner_searching_2007}
comprising \uwsize\ ill-typed programs.
%
The second is a new \ucsdbench\ data set, comprising \ucsdsize\
ill-typed programs.
%
We show that for both data sets, our technique is able to generate
witnesses for nearly 90\% of the programs, in under a second in the
vast majority of cases.
%
Furthermore, we show that a simple interactive strategy yields
compact counterexample traces with at most 5 steps for 57\%
of the programs, and at most 10 steps for 81\% of the programs.

>>>>

Type errors are a common stumbling block for students
trying to learn typed functional languages like \ocaml\
and \haskell.
%
Consider the ill-typed @fac@ function on the left in
Figure~\ref{fig:factorial}.
%
The function returns @true@ in the base case (instead of @1@),
and so \ocaml responds with the error message:
%
\begin{verbatim}
 File "fac.ml", line 5, characters 8-17:
 Error: This expression has type bool
        but an expression was expected
        of type int
\end{verbatim}
%
This message makes perfect sense to an expert who is familiar
with the language and has a good mental model of how the type
system works.
%
However, it may perplex a novice who has yet to develop such a
mental model.
%
To make matters worse, unification-based type inference algorithms
often report errors far removed from their source.
%
This further increases the novice's confusion and can actively mislead
them to focus their investigation on an irrelevant piece of code.
%
Much recent work has focused on analyzing unification constraints
to properly \emph{localize} a type error~\cite{Lerner2007-dt,Chen2014-gd,Zhang2014-lv,Pavlinovic2014-mr},
but an accurate source location still does not explain \emph{why} the
program is wrong.

\begin{figure}[t]
\centering
\begin{minipage}{.49\linewidth}
\centering
\begin{code}
  let rec fac n =
    if n <= 0 then
      true
    else
      n * fac (n-1)
\end{code}
\vspace{3em}
% \end{minipage}
% % \hspace{0.3in}
% \begin{minipage}{.39\linewidth}
\centering
  \includegraphics[height=1.5in]{fac-overview.png}
\end{minipage}
\begin{minipage}{.49\linewidth}
\centering
\includegraphics[height=3in]{fac-long.png}
\end{minipage}
\caption{(top-left) An ill-typed \texttt{fac} function; (bottom-left) Dynamically witnessing the type error in \texttt{fac}, showing only function calls; (right) The same trace, fully expanded.}
\label{fig:factorial}
\end{figure}

@@

\paragraph{3. Evaluating Witnesses}
%
Of course, the problem of finding witnesses is
undecidable. In fact, due to the necessarily
conservative nature of static typing, there
may not even exist any witnesses for a given
ill-typed program.
%
Thus, our approach is a heuristic that is only useful
if it can find \emph{compact} witnesses for
\emph{real-world} programs.
%
Our third contribution is an extensive evaluation of our approach
on two different sets of ill-typed programs obtained by instrumenting
compilers used in beginner's classes~(\S~\ref{sec:evaluation}).
%
The first is the \uwbench\ data set~\cite{Lerner2007-dt}
comprising \uwsize\ ill-typed programs.
%
The second is a new \ucsdbench\ data set, comprising \ucsdsize\
ill-typed programs.
%
We show that for both data sets, our technique is able to generate
witnesses for nearly 90\% of the programs, in under a second in the
vast majority of cases.
%
Furthermore, we show that a simple interactive strategy yields
compact counterexample traces with at most 5 steps for 57\%
of the programs, and at most 10 steps for 81\% of the programs.

### 2016-05-24 paper/icfp16-cameraready/overview.tex
misc

To defer input generation, we borrow the notion of a ``hole'' from
SmallCheck~\cite{runciman_smallcheck_2008}.
%
A hole --- written \vhole{\thole} --- is a \emph{placeholder} for a
value \ehole of some unknown type \thole.
%
We leave all inputs as uninstantiated holes until they are demanded by
the program, \eg due to a primitive operation like the @<=@ test.

@@

\subsection{Visualizing Witnesses}
\label{sec:visual-witness}
We have shown how to reliably find witnesses to type errors in \ocaml,
but this does not fully address our original goal --- to \emph{explain}
the errors.
%
Having identified an input vector that triggers a crash, a common next
step is to step through the program with a \emph{debugger} to observe
how the program evolves.
%
The existing debuggers and interpreters for \ocaml\ assume a type-correct
program, so unfortunately we cannot use them off-the-shelf.
%
Instead we extend our search for witnesses to produce an execution
trace.

>>>>

To defer input generation, we borrow the notion of a ``hole'' from
SmallCheck~\cite{Runciman2008-ka}.
%
A hole --- written \vhole{\thole} --- is a \emph{placeholder} for a
value \ehole of some unknown type \thole.
%
We leave all inputs as uninstantiated holes until they are demanded by
the program, \eg due to a primitive operation like the @<=@ test.

@@

\subsection{Visualizing Witnesses}
\label{sec:visual-witness}
We have described how to reliably find witnesses to type errors in \ocaml,
but this does not fully address our original goal --- to \emph{explain}
the errors.
%
Having identified an input vector that triggers a crash, a common next
step is to step through the program with a \emph{debugger} to observe
how the program evolves.
%
The existing debuggers and interpreters for \ocaml\ assume a type-correct
program, so unfortunately we cannot use them off-the-shelf.
%
Instead we extend our search for witnesses to produce an execution
trace.
